---
title: "p8105_hw2_lq2250"
author: "Lanlan Qing"
date: "2025-09-29"
output: github_document
---

```{r}
library(tidyverse)
```

## Problem 1

### 1. Clean the data in pols-months.csv
```{r}
pols_clean = read.csv("data_538/pols-month.csv", na = c("NA",".","")) |>
  janitor::clean_names() |>
  separate(mon, into = c("year", "month", "day"), sep = "-") |>
  mutate(month =
           case_match(
             month,
             "01" ~ "Jan", 
             "02" ~ "Feb", 
             "03" ~ "Mar", 
             "04" ~ "Apr",
             "05" ~ "May", 
             "06" ~ "Jun", 
             "07" ~ "Jul", 
             "08" ~ "Aug", 
             "09" ~ "Sep", 
             "10" ~ "Oct", 
             "11" ~ "Nov", 
             "12" ~ "Dec")
           ) |>
  mutate(
    president = if_else(prez_gop == 1, "gop","dem")) |>
  select(-prez_dem, -prez_gop, -day)|>
  arrange(year)
```


### 2. Clean the data in pols-months.csv
```{r}
snp_clean = read.csv("data_538/snp.csv", na = c("NA",".","")) |>
  separate(date, into = c("month", "day", "year"), sep = "/") |>
  mutate(month = 
           case_match( month, 
                       "1" ~ "Jan",
                       "2" ~ "Feb", 
                       "3" ~ "Mar",
                       "4" ~ "Apr", 
                       "5" ~ "May",
                       "6" ~ "Jun", 
                       "7" ~ "Jul", 
                       "8" ~ "Aug", 
                       "9" ~ "Sep", 
                       "10" ~ "Oct", 
                       "11" ~ "Nov", 
                       "12" ~ "Dec")
           ) |>
  arrange(year) |>
  select(year, month, close, -day)
```

### 3. Clean the data in unemployment.csv
```{r}
unemploy_read = read.csv("data_538/unemployment.csv", na = c("NA",".",""))

unemploy_clean = 
  pivot_longer(
    unemploy_read,
    Jan:Dec,
    names_to = "month",
    values_to = "percentage"
  )|>
  rename(year = Year) |>
  mutate(year = as.character(year))
```

```{r}
merge_pols_snp = left_join(pols_clean, snp_clean)
merge_df = left_join(merge_pols_snp, unemploy_clean)
```

**Interpretation** \
For the **original(uncleaned) datasets**, the dimension of pols-month dataset is 822x9; snp dataset is 11x2; unemployment dataset is of 787x2.

For all three datasets, they all have **the data of date(where later cleaned as "year", "month" and "day)**, but can be characterized as follows: 

 - the **pols-months** is related to the number of national politicians (president, governors, senators, representatives) who are democratic or republican at any given **date**
 
 - the **snp** is related to Standard & Poor’s stock market index (S&P), showing the closing values of the S&P stock index on the associated **date**
 
 - the **unemployment** is related to the percentage of unemployment on the associated **month of the year**
 
While and after **cleaning the data**:

 - we removed *day* column from both "pols-months" and "snp" dataset, whereas no *day* exists in the "unemployment" dataset. By removing *day*, three tables can be better joined together.
 
 - for "snp" dataset, only data from **1950-2015** year are recorded; "unemployment" was recorded from **1948-2015**; pols" was recorded from **1947-2015**. All missing data from 1947-1949 in "snp" and "unemployment" were indicated as "NA" in the resulting dataset.
 
 - we transferred the columns of *Jan* - *Dec* in **unemployment** under *month* column for year-month mergability of three tables. 



