---
title: "p8105_hw2_lq2250"
author: "Lanlan Qing"
date: "2025-09-29"
output: github_document
---

```{r message = FALSE}
library(tidyverse)
library(readxl)
library(lubridate)
library(stringr)
library(dplyr)
```

## Problem 1

### 1. Clean the data in pols-months.csv

```{r}
pols_clean = read_csv("data/pols-month.csv", na = c("NA",".","")) |>
  janitor::clean_names() |>
  separate(mon, into = c("year", "month", "day"), sep = "-") |>
  mutate(month =
           case_match(
             month,
             "01" ~ "Jan", 
             "02" ~ "Feb", 
             "03" ~ "Mar", 
             "04" ~ "Apr",
             "05" ~ "May", 
             "06" ~ "Jun", 
             "07" ~ "Jul", 
             "08" ~ "Aug", 
             "09" ~ "Sep", 
             "10" ~ "Oct", 
             "11" ~ "Nov", 
             "12" ~ "Dec")
           ) |>
  mutate(
    president = if_else(prez_gop == 1, "gop","dem")) |>
  select(-prez_dem, -prez_gop, -day)|>
  arrange(year)
```

### 2. Clean the data in pols-months.csv

```{r}
snp_clean = read_csv("data/snp.csv", na = c("NA",".","")) |>
  separate(date, into = c("month", "day", "year"), sep = "/") |>
  mutate(month = 
           case_match( month, 
                       "1" ~ "Jan",
                       "2" ~ "Feb", 
                       "3" ~ "Mar",
                       "4" ~ "Apr", 
                       "5" ~ "May",
                       "6" ~ "Jun", 
                       "7" ~ "Jul", 
                       "8" ~ "Aug", 
                       "9" ~ "Sep", 
                       "10" ~ "Oct", 
                       "11" ~ "Nov", 
                       "12" ~ "Dec")
           ) |>
  arrange(year) |>
  select(year, month, close, -day)
```

### 3. Clean the data in unemployment.csv

```{r}
unemploy_read = read_csv("data/unemployment.csv", na = c("NA",".",""))

unemploy_clean = 
  pivot_longer(
    unemploy_read,
    Jan:Dec,
    names_to = "month",
    values_to = "percentage"
  )|>
  rename(year = Year) |>
  mutate(year = as.character(year))
```

```{r}
merge_pols_snp = left_join(pols_clean, snp_clean)
merge_df = left_join(merge_pols_snp, unemploy_clean)
```

**Interpretation**\
For the **original(uncleaned) datasets**, the dimension of pols-month dataset is 822x9; snp dataset is 11x2; unemployment dataset is of 787x2.

For all three datasets, they all have **the data of date(where later cleaned as "year", "month" and "day)**, but can be characterized as follows:

-   the **pols-months** is related to the number of national politicians (president, governors, senators, representatives) who are democratic or republican at any given **date**

-   the **snp** is related to Standard & Poor’s stock market index (S&P), showing the closing values of the S&P stock index on the associated **date**

-   the **unemployment** is related to the percentage of unemployment on the associated **month of the year**

While and after **cleaning the data**:

-   we removed *day* column from both "pols-months" and "snp" dataset, whereas no *day* exists in the "unemployment" dataset. By removing *day*, three tables can be better joined together.

-   for "snp" dataset, only data from 1950-2015 year are recorded; "unemployment" was recorded from 1948-2015; pols" was recorded from 1947-2015. All missing data from 1947-1949 in "snp" and "unemployment" were indicated as "NA" in the resulting dataset.

-   we transferred the columns of *Jan* - *Dec* in **unemployment** under *month* column for year-month mergability of three tables.

-   the cleaned datasets' dimensions are: pols : `r nrow(pols_clean)`x`r ncol(pols_clean)`; snp : `r nrow(snp_clean)`x `r ncol(snp_clean)`; unemployment : `r nrow(unemploy_clean)`x `r ncol(unemploy_clean)`

## Problem 2

### 1. Import and clean the data for Mr. Trash Wheel

```{r}
mr_xl = read_excel("data/trash_wheel.xlsx",
           sheet = "Mr. Trash Wheel",
           na = c("NA", ".",""),
           range = "A2:N709") |>
  janitor::clean_names() |>
  rename(weight = weight_tons,
         volume = volume_cubic_yards) |>
  drop_na(dumpster) |>
  mutate(sports_balls = as.integer(round(sports_balls))) |>
  mutate(type = "mr") |>
  select(-year, -month) |>
  select(type, dumpster, date, everything())
  
```

### 2. Import and clean the data for Professor Trash Wheel

```{r}
prof_xl = read_excel("data/trash_wheel.xlsx",
           sheet = "Professor Trash Wheel",
           na = c("NA", ".",""),
           range = "A2:M134") |>
  janitor::clean_names() |>
  rename(weight = weight_tons,
         volume = volume_cubic_yards) |>
  filter(!is.na(dumpster)) |>
  mutate(sports_balls = NA) |>
  mutate(type = "prof") |>
  select(-year, -month) |>
  select(type, dumpster, date, everything())
  
```

### 3. Import and clean the data for Gwynnda Trash Wheel

```{r}
gwy_xl = read_excel("data/trash_wheel.xlsx",
           sheet = "Gwynns Falls Trash Wheel",
           na = c("NA", ".",""),
           range = "A2:L351") |>
  janitor::clean_names() |>
  rename(weight = weight_tons,
         volume = volume_cubic_yards) |>
  filter(!is.na(dumpster))|>
  mutate(glass_bottles = NA,
         sports_balls = NA) |>
  mutate(type = "gwy") |>
  select(-year, -month) |>
  select(type, dumpster, date, everything())
  
```

### 4. Combine 3 tables

```{r}
combine_xl = bind_rows(prof_xl, gwy_xl, mr_xl) |>
  separate(date, into = c("year", "month", "day"), sep = "-")
```

### 5. Interpretation

-   The total number of observations is `r nrow(combine_xl)`.

-   The key variables are `r colnames(combine_xl)`.

-   The total weight collected by Professor Trash Wheel is: `r combine_xl |> filter(type == "prof") |> summarise(total = sum(weight, na.rm = TRUE)) |> pull(total)`

-   The total number of cigarette butts collected by Gwynns in June of 2022 is: `r combine_xl |> filter((year == "2022") & (month == "06")) |> summarise(total = sum(cigarette_butts, na.rm = TRUE)) |> pull(total)`

## Problem 3

### 1.Read and clean the zip code dataset

```{r}
zip_codes = read_csv("zillow_data/Zip Codes.csv", na = c("NA", ".", "")) |>
  janitor::clean_names() |>
  mutate(file_date = mdy(file_date)) |>
  select(-state_fips, -county_code) |>
  mutate(state = "NY", city = "New York") |>
  relocate(file_date, state, city, county, zip_code)
```

### 2.Read and clean the zip rental price dataset

```{r}
zip_price_read = read_csv("zillow_data/Zip_zori_uc_sfrcondomfr_sm_month_NYC.csv", 
                          na = c("NA", ".", ""))
  
zip_price = pivot_longer(
  zip_price_read,
  `2015-01-31`:`2024-08-31`,
  names_to = "file_date",
  values_to = "rental_price") |>
  janitor::clean_names() |>
  rename(zip_code = region_name) |>
  select(-region_type, -state_name) |>
  rename(county = county_name) |>
  mutate(county = str_remove(county, " County")) |>
  relocate(file_date, state, city, county, zip_code) |>
  mutate(file_date = 
           ymd(file_date))
  
```

### 3.Join two datasets

```{r}
final_df = full_join(zip_codes, zip_price) |>
  relocate(rental_price, .after = county_fips)
```

### 4.Interpretation of resulting tidy dataset

-   total observations: `r nrow(final_df)`

-   there are `r length(unique(final_df$zip_code))` unique zip codes

-   there are `r length(unique(na.omit((final_df$neighborhood))))` unique neighborhoods

-   zip codes appear in the ZIP code dataset but not in the Zillow Rental Price are: `r anti_join(distinct(zip_codes, .keep_all = TRUE), distinct(zip_price, .keep_all = TRUE), by = "zip_code") |> pull(zip_code) |> format(scientific = FALSE) |> sort()`. The reasons why these ZIP codes might be excluded can be attributable to the time period recorded differently, where ZIP codes collected zip codes in 2007 year and zillow rental price collected data from 2015-2024. There is a 8-years-gap, which means that some of the places might be demolished or relocated.

### 5. Rental price comparisons

```{r}

rental_2020 = filter(final_df, ((year(file_date) == 2020) & (month(file_date) == 01))) |>
  drop_na(rental_price) |>
  mutate(price_2020 = rental_price) |>
select(zip_code,county,neighborhood, price_2020)

rental_2021 = filter(final_df, ((year(file_date) == 2021) & (month(file_date) == 01))) |>
  drop_na(rental_price) |>
  mutate(price_2021 = rental_price) |>
select(zip_code,county,neighborhood, price_2021)

# Make a table that is composed of 10 largest drop of rental price from 2020-01 to 2021-01
rental_dif_data = left_join(rental_2020, rental_2021) |>
  mutate(rental_dif = price_2021 - price_2020) |>
  arrange(desc(rental_dif)) |>
  slice_tail(n = 10)

rental_compare_table <- rental_dif_data |>
  knitr::kable()
```

According to the table, the 10 ZIP codes are: `r format(rental_dif_data$zip_code,science = FALSE)`.
